---
title: "Introduction to Data Science in Health and Social Care"
subtitle: "Week 5"
runtime: shiny
output: html_document
---

<!-- Press "Run Document" in RStudio to start (or CMD + SHIFT + K or CTRL + SHIFT + K) -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install and load require packages
list.of.packages <- c("tidyverse", "rmarkdown", "gt", "shiny")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[, "Package"])]
if(length(new.packages)){install.packages(new.packages)}

lapply(list.of.packages, require, character.only = TRUE)

# define cars_df using mtcars
cars_df <- mtcars %>%
  # add car and make columns
  mutate(
    car  = rownames(mtcars),
    make = word(car, 1)
    ) %>%
  relocate(car, make) %>%
  # define data types
  mutate(
    cyl  = as.factor(cyl),
    vs   = as.factor(vs),
    am   = as.factor(am),
    gear = as.factor(gear),
    carb = as.factor(carb)
  )
  
```


## Our Data

We are using the `mtcars` (Motor Trend Car Road Tests) dataset which includes information about 32 cars. The `cars_df` is a dataframe with a few extra columns, it is already loaded for your convince. If you are interested how this dataframe was formed, you can see the code in lines 20 - 35.

The dataframe contains many types of variables, qualitative and quantitative, categorical (discrete) and continuous.

The columns:

| Column Name | Description                              |
| ----------- | ---------------------------------------- |
| car         | Car name                                 |
| make        | Car manufacturer                         |
| mpg         | Miles/(US) gallon                        |
| cyl         | Number of cylinders                      |
| disp        | Displacement (cu.in.)                    |
| hp          | Gross horsepower                         |
| drat        | Rear axle ratio                          |
| wt          | Weight (1000 lbs)                        |
| qsec        | 1/4 mile time                            |
| vs          | Engine (0 = V-shaped, 1 = straight)      |
| am          | Transmission (0 = automatic, 1 = manual) |
| gear        | Number of forward gears                  |
| carb        | Number of carburetors                    |


The first few rows of `cars_df` are:

``` {r head-cars}
gt(head( cars_df ))
```

## Make Changes Here

Tasks:

1. You can select one (`x_variable`) or two (`x_variable` and `y_variable`) from the columns names of `cars_df`
2. See how the data, plots, and code has changed
3. Take a screenshot (CMD + SHIFT + 4 or WINDOWS + SHIFT + S) and tell a short story with the plot
4. Try another combination of column names

*If you only want to use one variable, set `y_variable` as null, i.e. `y_variable <- NULL`*

```{r inputs, echo=FALSE}
selectInput("x_variable", 
            label = "x_variable:",
            choices = names(cars_df), 
            selected = "cyl")

selectInput("y_variable", 
            label = "y_variable:",
            choices = c('NULL', names(cars_df)), 
            selected = "NULL")

```

## Display Data

The following code will display the data you have selected above. There is no need to change the following code.

The selected columns:

```{r, echo=FALSE}
renderPrint({
  if (input$y_variable == "NULL"){
    paste0("head( cars_df %>% select(", input$x_variable, ") )")
  } else {
    paste0("head( cars_df %>% select(",input$x_variable,", ", input$y_variable, ") )")
  }
})


``` 


```{r wrangle, echo=FALSE}
# Wrangle Data ----
wrangled_cars_df <- reactive({
  if (input$y_variable == "NULL"){
    cars_df %>%
      select(all_of(input$x_variable))
  } else {
    cars_df %>%
      select(all_of(c(input$x_variable, input$y_variable)))
  }
  })

# Define Data Class ----
x_class <- reactive({
  class(cars_df[, input$x_variable])
  })
y_class <- reactive({
  if(!is.null(input$y_variable)){
    y_class <- class(cars_df[, input$y_variable])
  } else {
    y_class <- NULL
  }
  y_class
  })
  

# display dataframe
renderTable(gt(head(wrangled_cars_df())))

```


```{r , echo=FALSE}
# Plots ----
renderPrint({
  if (!(input$y_variable %in% names(cars_df)) & (x_class() != "numeric")){
    ## Single Variable Categorical (Bar Plot)
    paste0("ggplot(wrangled_cars_df, aes(x = ", input$x_variable, ")) + geom_bar()") 

  } else if (!(input$y_variable %in% names(cars_df)) & (x_class() == "numeric")){
    ## Single Variable Continuous (Histogram Plot)
    paste0("ggplot(wrangled_cars_df, aes(x = ", input$x_variable, ")) + geom_histogram()") 
  
  } else if ((x_class() != "numeric") & (y_class() == "numeric")) {
    ## Two Variables Categorical-Continuous (Box Plot)
    paste0("ggplot(wrangled_cars_df, aes(x = ", input$x_variable, ", y = ", input$y_variable,")) + geom_boxplot()") 
  
  } else if ((x_class() == "numeric") & (y_class() != "numeric")) {
    ## Two Variables Continuous-Categorical (Box Plot)
    paste0("ggplot(wrangled_cars_df, aes(x = ", input$x_variable, ", y = ", input$y_variable,")) + geom_boxplot()") 
  
  } else if ((x_class() != "numeric") & (y_class() != "numeric")) {
    ## Two Variables Categorical-Categorical (2D Histogram)
    paste0("ggplot(wrangled_cars_df, aes(x = ", input$x_variable, ", y = ", input$y_variable,")) + geom_bin2d()") 
  
  } else if ((x_class() == "numeric") & (y_class() == "numeric")) {
    ## Two Variables Continuous-Continuous (Scatter Plot)
    paste0("ggplot(wrangled_cars_df, aes(x = ", input$x_variable, ", y = ", input$y_variable,")) + geom_point()") 
  
  }
  
})


```

```{r plot, echo=FALSE}
# Plots ----
renderPlot({
  if (!(input$y_variable %in% names(cars_df)) & (x_class() != "numeric")){
    ## Single Variable Categorical (Bar Plot)
    
    cars_plot <- ggplot(
      wrangled_cars_df(),
      aes_string(
        x    = input$x_variable,
        fill = input$x_variable
      )) +
      geom_bar() +
      theme(legend.position = "none")
  
    # rotate labels if too many options in the x axis
    if (length(unique(wrangled_cars_df()[,input$x_variable])) > 10){
      cars_plot <- cars_plot +
        theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))
    }
  
  
  } else if (!(input$y_variable %in% names(cars_df)) & (x_class() == "numeric")){
    ## Single Variable Continuous (Histogram Plot)
    cars_plot <- ggplot(
      wrangled_cars_df(),
      aes_string(
        x = input$x_variable
      )) +
      geom_histogram(bins = 10)
  
  
  } else if ((x_class() != "numeric") & (y_class() == "numeric")) {
    ## Two Variables Categorical-Continuous (Box Plot)
    cars_plot <- ggplot(
      wrangled_cars_df(),
      aes_string(
        x   = input$x_variable,
        y   = input$y_variable,
        col = input$x_variable
      )) +
      geom_boxplot() +
      theme(legend.position = "none")
  
    # rotate labels if too many options in the x axis
    if (length(unique(wrangled_cars_df()[,input$x_variable])) > 10){
      cars_plot <- cars_plot +
        theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))
    }
  
  } else if ((x_class() == "numeric") & (y_class() != "numeric")) {
    ## Two Variables Continuous-Categorical (Box Plot)
    cars_plot <- ggplot(
      wrangled_cars_df(),
      aes_string(
        x   = input$x_variable,
        y   = input$y_variable,
        col = input$y_variable
      )) +
      geom_boxplot() +
      theme(legend.position = "none")
  
  
  } else if ((x_class() != "numeric") & (y_class() != "numeric")) {
    ## Two Variables Categorical-Categorical (2D Histogram)
    cars_plot <- ggplot(
      wrangled_cars_df(),
      aes_string(
        x = input$x_variable,
        y = input$y_variable
      )) +
      geom_bin2d()
  
  } else if ((x_class() == "numeric") & (y_class() == "numeric")) {
    ## Two Variables Continuous-Continuous (Scatter Plot)
    cars_plot <- ggplot(
      wrangled_cars_df(),
      aes_string(
        x = input$x_variable,
        y = input$y_variable
      )) +
      geom_point(alpha = 0.8)
  
  }
  
  # display plot
  cars_plot
})


```

----

*Kevin Tsang (2022-08-12)*