---
title: "Week 8 Tutorial: TBI Data SOLUTIONS"
date: "`r Sys.Date()`"
output: 
  html_document: 
    theme: paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You will need to load the required libraries first: `tidyverse`, `here`, `knitr`, and `kableExtra`.

```{r libraries, message = FALSE}
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
```

## The Data: TBI 

We will be using traumatic brain injury (TBI) data from a Tidy Tuesday challenge. You can learn more about [the challenge and the data here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-03-24/readme.md) (spend 3-5 minutes looking at the linked page). Be sure to read over the data dictionary to make sure you understand what variables are included in the datasets and what they measure. 

In the following tasks we will work through a data storytelling data flow. Imagine that you are preparing a brief for the local health authority, and they have asked you to answer the following questions. For each question, write a chunk of code that would provide the answer in a clean and reproducible way.

Let's start by loading in the data, which is available online:

```{r loading-data, message = FALSE}
tbi_age <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_age.csv")

tbi_year <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_year.csv")
```

Last week we learned that as part of good coding practice, your folder structure plays a key role. Now, set up folders in your working directory: raw_data, scripts, plots, and tables. Put this RMarkdown document in your scripts folder. 

If you are working with Projects (which is recommended), be sure that these folders are accessible by your R Project. You can do this by (a) creating a new project or (b) creating your folder structure within an existing R Project space. 

## Tasks {.tabset}
### Task 1

Save both datasets in the raw_data folder (use the 'here' library). Particularly when using data for an analysis that is available online and continually updated, it is a good idea to save a version of the raw data that you used. 

```{r task-1}
write_csv(tbi_age, file = here("raw_data", "tbi_age.csv"))
write_csv(tbi_year, file = here("raw_data", "tbi_year.csv"))
```


### Task 2.1

Use your favourite functions for eyeballing data and write a couple of sentences about each dataset (its dimensions, the variables inside, types of data, etc.)

```{r task-2-1, eval=FALSE}
glimpse(tbi_age)
summary(tbi_age)
str(tbi_age)

glimpse(tbi_year)
summary(tbi_year)
str(tbi_year)
```

The `tbi_age` dataset has 231 rows and 5 columns. It includes 5 variables: age group, type of outcome, injury mechanism, estimated number of injuries in total, and estimated rate per 100000 population. Estimated number and rate are numeric, type of outcome and injury mechanism, are categorical nominal variables stored as strings, and age is a categorical ordinal variable, also stored as a string/character variable.

The `tbi_year` dataset 216 rows and 5 columns. It includes 5 variables: injury mechanism, type of outcome, year, estimated number of injuries in total, and estimated rate per 100000 population. Year, estimated number and rate are numeric, type of outcome and injury mechanism, are categorical nominal variables stored as strings.

### Task 2.2

Are there any variables in the 2 datasets that should be treated as factors? Decide which ones, and turn them into factors. 

Once you've done that, think about the order of factor levels. Does it need to be changed? If so, change it.


```{r task-2-2}
tbi_age <- tbi_age %>% 
  mutate(type = factor(type),
         injury_mechanism = factor(injury_mechanism),
         age_group = factor(age_group, levels = c("0-17", "0-4", "5-14", "15-24",
                                                  "25-34", "35-44", "45-54", "55-64",
                                                  "65-74", "75+", "Total")))

tbi_year <- tbi_year %>% 
  mutate(injury_mechanism = factor(injury_mechanism),
         type = factor(type))
```


### Task 3

Which injury mechanism caused the most deaths in total, across age groups?

```{r task-3}
tbi_age %>% 
  filter(age_group == "Total" & type == "Deaths") %>% 
  arrange(desc(number_est)) %>% #arrange in descending order
  head(n = 1) # show only the first row (highest number)
```

### Task 4

Looking just at the totals across age groups, what is the number of ED visits, hospitalisations and deaths associated with each injury mechanism? Present the results as a table. Save this table in your tables folder.

Reminder, a great resource for all of the different options around formatting with `kable` tables and `kableExtra` can be [found here](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html#Table_Styles). I reference it pretty much every time I go to make a table! 

There are a few approaches that could be taken to answer this question, I have outlined 2 below: 

**Option 1** A good approach here is to pivot the table, so we get each injury type as a row, and each outcome as a column:

```{r task-4-approach-1, message = FALSE}

tbi_age %>% 
  filter(age_group == "Total") %>% 
  select(-rate_est, -age_group) %>% 
  pivot_wider(names_from = "type", values_from = "number_est")

# Now we can format this into a nice table 

total_injuries_table <- tbi_age %>% 
  filter(age_group == "Total") %>% 
  select(-rate_est, -age_group) %>% 
  pivot_wider(names_from = "type", values_from = "number_est") %>%
  kbl(col.names = c("Injury Mechanism",
                      "Emergency Department Visit", 
                      "Hospitalizations", 
                      "Deaths"),
        caption = "TBI: Injury Mechanism by Type of Measure") %>% 
  kable_classic_2(bootstrap_options = "striped", full_width = F) %>%
  footnote(general = "Data from CDC and TidyTuesday 2020-03-24",
           symbol = "Values represent estimated observed cases in 2014",
           footnote_as_chunk = T, 
           title_format = "italic")
  
total_injuries_table

# And now saving the table with the helpful save_kable() function

save_kable(total_injuries_table, file = here("tables", "total_injuries_table.pdf"))
```

**Option 2** rather than having each outcome as a column, a more simple approach would be to keep the outcomes as rows, though our table is a bit more messy as the type of measure is repeated 

```{r task-4-approach-2, message = FALSE}

tbi_age %>% 
  filter(age_group == "Total") %>% 
  select(-rate_est, -age_group) %>% 
  arrange(desc(number_est))

# Now we can format this into a nice table 

total_injuries_table_v2 <- tbi_age %>% 
  filter(age_group == "Total") %>% 
  select(-rate_est, -age_group) %>% 
  arrange(desc(number_est)) %>% 
  kbl(col.names = c("Type", 
                      "Injury Mechanism",
                      "Estimated observed cases in 2014"),
        caption = "TBI: Injury Mechanism and Type of Measure") %>% 
  kable_classic(full_width = F, html_font = "Arial Narrow") %>% 
  footnote(general = "Data from CDC and TidyTuesday 2020-03-24",
           footnote_as_chunk = T, 
           title_format = "italic")
  
total_injuries_table_v2

save_kable(total_injuries_table_v2, file = here("tables", "total_injuries_table_v2.pdf"))
```


### Task 5

Focusing on the oldest age group (75+), draw a bar chart illustrating the number of hospitalisations associated with each injury mechanism. Make sure that you add a title and appropriate axis labels. Save this chart in your plots folder.

*Note* the labels are very long, and there are different ways to deal with that:
https://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2

I have outlined 3 potential options (which could be combined), though there are more! 

**Option 1**: `str_wrap()` to add a line break after every 15 characters 

```{r task-5-approach-1, message=FALSE}
plot1_v1 <- tbi_age %>% 
  filter(age_group == "75+" & type == "Hospitalizations") %>% 
  # the str_wrap function from stringr adds line breaks in our injury_mechanism
  # labels, so they look better in the plot
  mutate(injury_mechanism = str_wrap(injury_mechanism, width = 15)) %>% 
  # reorder will sort the bars by height, so that the injury mechanism
  # with most cases is first
  ggplot(aes(x = reorder(injury_mechanism, desc(-number_est)), y = number_est)) +
  geom_col(fill = "steelblue4") +  
  labs(title = "Injuries, by mechanism, in the 75+ age group",
       x = "Injury mechanism", y = "Number of injuries") + 
  theme_bw()

plot1_v1 

ggsave(here("plots", "Injuries_oldest_group_1.pdf"))
# ggsave saves the most recent plot by default, but you can specify the plot with the 'plot' argument
```


**Option 2**: flip the coordinates so the labels do not overlap 

```{r task-5-approach-2, message=FALSE}
plot1_v2 <- tbi_age %>% 
  filter(age_group == "75+" & type == "Hospitalizations") %>% 
  ggplot(aes(x = reorder(injury_mechanism, desc(-number_est)), y = number_est)) +
  geom_col(fill = "steelblue4") +  
  labs(title = "Injuries, by mechanism, in the 75+ age group",
       x = "Injury mechanism", y = "Number of injuries") + 
  theme_bw() + 
  coord_flip() # this swaps x and y coordinates, making the labels more visible

plot1_v2

ggsave(here("plots", "Injuries_oldest_group_2.pdf"))
```

**Option 3**: rotate the x-axis labels so they do not overlap


```{r task-5-approach-3, message=FALSE}
plot1_v3 <- tbi_age %>% 
  filter(age_group == "75+" & type == "Hospitalizations") %>% 
  ggplot(aes(x = reorder(injury_mechanism, -number_est), y = number_est)) +
  geom_col(fill = "steelblue4") +  
  scale_x_discrete(guide = guide_axis(angle = 70)) +
  labs(title = "Injuries, by mechanism, in the 75+ age group",
       x = "Injury mechanism", y = "Number of injuries") + 
  theme_bw()

plot1_v3

ggsave(here("plots", "Injuries_oldest_group_3.pdf"))
```

**Option 4**: We could also combine these approaches (specifically shown below is option 1 and 2)

```{r task-5-combo, message=FALSE}
plot1_v4 <- tbi_age %>% 
  filter(age_group == "75+" & type == "Hospitalizations") %>% 
  mutate(injury_mechanism = str_wrap(injury_mechanism, width = 15)) %>% 
  ggplot(aes(x = reorder(injury_mechanism, desc(-number_est)), y = number_est)) +
  geom_col(fill = "steelblue4") +  
  labs(title = "Injuries, by mechanism, in the 75+ age group",
       x = "Injury mechanism", y = "Number of injuries") + 
  theme_bw() + 
  coord_flip()

plot1_v4

ggsave(here("plots", "Injuries_oldest_group_4.pdf"))
```


### Task 6

Plot a line chart that depicts the rate of deaths, per 100000 population, 
from traumatic brain injury in years 2006-2014. Which datset do you think we will need to use for this? Save this chart in your plots folder.

```{r task-6, message=FALSE}
plot2 <- tbi_year %>% 
  filter(injury_mechanism == "Total" & type == "Deaths") %>% 
  ggplot(aes(x = year, y = rate_est)) +
  geom_line() + 
  geom_point() + # using two geoms makes the data points more visible
  scale_x_continuous(breaks = seq(2006, 2014, 1)) + # to have each year shown on the x-axis
  labs(title = "Rate of deaths (per 100000 population) from traumatic brain injury",
       x = "Year", 
       y = "Estimated rate") + 
  theme_bw()

plot2

ggsave(here("plots", "Total_death_rates.pdf"))
```

### Task 7

The CDC produced a report, entitled the TBI Surveillance Report, which you can find at [this URL](https://www.cdc.gov/traumaticbraininjury/pdf/TBI-Surveillance-Report-FINAL_508.pdf). 

On page 14 of the report (page 15 of the pdf document), you'll find Figure 2A, showing the rates, per 100000 population, of TBI-related ED visits, by year, and mechanism of injury, 2006-2014. Try and re-create this figure.

First, focus just on depicting the data (you can use the default settings of ggplot). If you still have time, try reproducing this figure as closely as possible (colors, style, etc.). Save the final version of your figure in your plots folder.

```{r task-7, fig.height=8.5, fig.width=10, warning=FALSE, message=FALSE}

plot3 <- tbi_year %>% 
  filter(type == "Emergency Department Visit" & injury_mechanism != "Total") %>% 
  ggplot(aes(x = year, y = rate_est, colour = injury_mechanism)) +
  geom_line(size = 0.8) + 
  geom_point() +
  ylim(0, 400) +  
  labs(title = "FIGURE 2A: ESTIMATED AGE-ADJUSTED RATES, PER 100,000 POPULATION,OF TRAUMATIC BRAIN INJURY–\nRELATED EMERGENCY DEPARTMENT VISITS, BY YEAR AND MECHANISM OF INJURY, 2006-2014",
       y = "Age-adjusted rates of TBI-related ED visits", 
       x = NULL) +
  scale_x_continuous(breaks = seq(2006, 2014, 1)) + 
  scale_color_manual(values = c("#ed9338", "#fec836", "#7d3484", "#16928a", 
                                "#8ab9d2", "#1f6db2", "#16716e")) + 
  theme_bw() + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        plot.title = element_text(face = "bold", hjust = 0.5), #hjust centers the title 
        #the value of hjust and vjust are only defined between 0 and 1: 0 = left-justified & 1 = right-justified
        axis.line = element_line(colour = "black"),
        axis.title = element_text(face = "bold"),
        axis.text.x = element_text(face = "bold", color = "#000000"), 
        #instead of the hex code for black, you could also use the default color "black" as in line 307 
        axis.text.y = element_text(face = "bold", color = "#000000"),
        legend.position = "bottom",
        legend.title = element_blank()) 

plot3 

ggsave(here("plots", "TBI_related_ED_visits.pdf"), width = 11, height = 8.5)
```

**Note:** to get the exact color palette used in the report, I used the Chrome eye dropper extension to read the specific colors in the plot. From there, I took the HEX codes and put them into my ggplot code using the `scale_color_manual()` function as I wanted to manually tell R the color palette to use. This is largely unnecessary, but wanted to emphasis that I *did not* just happen to know the HEX codes for the colors. 

# {-}

---
*Brittany Blankinship & Kasia Banas (2022)* 

