---
title: "Week 9 Risk Prediction"
author: "Kevin Tsang"
date: "2022-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(PRROC)
```

## Introduction

Data source: https://www.opendata.nhs.scot/dataset/unintentional-injuries

Research question: Given someone (knowing their age) has died from injury, what probability that it was from a "fall"?

```{r functions}
# functions ----
best_gma <- function(roc){
  # Function to calculate the best threshold based on geometric mean accuracy (GMA)
  spe <- 1 - roc$curve[,1]
  sen <- roc$curve[,2]
  
  thresh <- roc$curve[,3]
  
  gma <- sqrt(spe * sen)
  
  opt_gma    <- max(gma)
  opt_thresh <- thresh[which.max(gma)]
  opt_spe    <- spe[which.max(gma)]
  opt_sen    <- sen[which.max(gma)]
  return(list(opt_gma    = opt_gma,
              opt_thresh = opt_thresh,
              opt_spe    = opt_spe,
              opt_sen    = opt_sen))
}
plot_spe_sen <- function(roc){
  # Plot the specificity and sensitivity across different thresholds
  # Note: this is not the ROC curve
  plot_df <- data.frame(
    spe    = 1 - roc$curve[,1],
    sen    = roc$curve[,2],
    thresh = roc$curve[,3])
  
  plot_df <- plot_df %>%
    pivot_longer(cols      = c("spe", "sen"), 
                 names_to  = "metric",
                 values_to = "value")
  
  spe_sen_plot <- ggplot(plot_df,
                         aes(x   = thresh,
                             y   = value,
                             col = metric)) +
    geom_step()
  
  print(spe_sen_plot)
  return(spe_sen_plot)
}
```

```{r import-data}
# use either a local data file import or URL import
deaths_raw <- read_csv("ui_deaths_2022.csv")
# deaths_raw <- read_csv("https://www.opendata.nhs.scot/dataset/b0135993-3d8a-4f3b-afcf-e01f4d52137c/resource/89807e07-fc5f-4b5e-a077-e4cf59491139/download/ui_deaths_-2022-.csv")

# remove aggregate values
deaths <- deaths_raw %>%
  clean_names() %>%
  filter(age_group   != "All",
         sex         != "All",
         injury_type != "All")
```

## Data Wrangling

```{r}
deaths <- deaths %>%
  group_by(year, age_group, sex, injury_type) %>%
  summarise(numberof_deaths = sum(numberof_deaths, na.rm = TRUE)) %>%
  ungroup() %>%
  # re order factors
  mutate(injury_type = factor(injury_type),
         age_group = factor(age_group,
                            levels = c("0-4 years",
                                       "5-9 years",
                                       "10-14 years",
                                       "15-24 years",
                                       "25-44 years",
                                       "45-64 years",
                                       "65-74 years",
                                       "75plus years"))) %>%
  mutate(injury_type = fct_relevel(injury_type, "Other", after = Inf)) %>%
  # create numeric variable from ordinal data
  mutate(age_group_num = recode(age_group, 
                                "0-4 years"    = 0,
                                "5-9 years"    = 5,
                                "10-14 years"  = 10,
                                "15-24 years"  = 15,
                                "25-44 years"  = 25,
                                "45-64 years"  = 45,
                                "65-74 years"  = 65,
                                "75plus years" = 75)) %>%
  # one hot encoding categorical variable
  mutate(age_group0_4yo = ifelse(age_group    == "0-4 years", 1, 0),
         age_group5_9yo = ifelse(age_group    == "5-9 years", 1, 0),
         age_group10_14yo = ifelse(age_group  == "10-14 years", 1, 0),
         age_group15_24yo = ifelse(age_group  == "15-24 years", 1, 0),
         age_group25_54yo = ifelse(age_group  == "25-44 years", 1, 0),
         age_group45_64yo = ifelse(age_group  == "45-64 years", 1, 0),
         age_group65_74yo = ifelse(age_group  == "65-74 years", 1, 0),
         age_group75plusyo = ifelse(age_group == "75plus years", 1, 0)) %>%
  # create binary variables
  mutate(sex_male = ifelse(sex == "Male", 1, 0)) %>%
  mutate(falls = injury_type == "Falls")

# expand into 1 patient per row
deaths_patient <- deaths %>%
  uncount(numberof_deaths)
```

## Split data into training set, validation set, testing set

Testing set is data from 2021. Training and validation set is 80%-20% random split.

```{r}
set.seed(123)
# split train-test (year < 2021, year == 2021)
deaths_patient_falls_train_raw <- deaths_patient %>%
  filter(year != 2021)

# split train-valid (80%-20%)
train_valid_split <- sample(c(rep(TRUE,8), rep(FALSE,2)), 
                            size = nrow(deaths_patient_falls_train_raw), 
                            replace = TRUE)
deaths_patient_falls_train <- deaths_patient_falls_train_raw[train_valid_split,]
deaths_patient_falls_valid <- deaths_patient_falls_train_raw[!train_valid_split,]
deaths_patient_falls_test  <- deaths_patient %>%
  filter(year == 2021)

```

## Plots about the population

```{r}
age_fall_plot <- deaths %>% 
  filter(injury_type == "Falls",
         year != 2021) %>% 
  group_by(age_group, year) %>%
  summarise(numberof_deaths = sum(numberof_deaths, na.rm = TRUE)) %>%
  ggplot(
    aes(x   = age_group,
        y   = numberof_deaths,
        col = as.factor(year))) +
  geom_point() + 
  labs(
    x       = "Age Group",
    y       = "Number of Deaths",
    title   = "Deaths from Falls in Scotland (2012 - 2020)",
    col     = "Year",
    caption = "Source: Public Health Scotland"
  ) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
age_fall_plot

# ggsave("deaths_age_fall_plot.png", age_fall_plot,
#        width = 1500, height = 1200, units = "px")

```

``` {r}
death_type_year_plot <- deaths %>%
  filter(year != 2021) %>%
  group_by(year) %>%
  mutate(annual_deaths = sum(numberof_deaths, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(numberof_deaths_perc = numberof_deaths / annual_deaths * 100) %>%
  group_by(injury_type, year) %>%
  summarise(numberof_deaths      = sum(numberof_deaths,      na.rm = TRUE),
            numberof_deaths_perc = sum(numberof_deaths_perc, na.rm = TRUE)) %>%
  ggplot(
    aes(x    = year,
        y    = numberof_deaths_perc,
        fill = injury_type)) +
  geom_col() +
  labs(
    x        = "Year",
    y        = "Number of Deaths per Year (%)",
    title    = "Proportion of Deaths by Different Injury Types",
    subtitle = "(2012 - 2020)",
    fill     = "Injury Type",
    caption  = "Source: Public Health Scotland"
  ) +
  theme_light()
death_type_year_plot

# ggsave("deaths_type_year_plot.png", death_type_year_plot,
#        width = 1500, height = 1200, units = "px")
```

``` {r}
age_fall_proportion_plot <- deaths %>% 
  filter(year != 2021) %>%
  group_by(age_group) %>%
  mutate(age_group_deaths = sum(numberof_deaths, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(numberof_deaths_perc = numberof_deaths / age_group_deaths * 100) %>%
  group_by(falls, age_group) %>%
  summarise(
    numberof_deaths      = sum(numberof_deaths,      na.rm = TRUE),
    numberof_deaths_perc = sum(numberof_deaths_perc, na.rm = TRUE)
    ) %>%
  ggplot(
    aes(x    = age_group,
        y    = numberof_deaths_perc,
        fill = falls)) +
  geom_col() + 
  labs(
    x        = "Age Group",
    y        = "Number of Deaths per Age Group (%)",
    title    = "Proportion of Deaths from Falls in Scotland",
    subtitle = "(2012 - 2020)",
    fill     = "Injury Type",
    caption  = "Source: Public Health Scotland"
  ) +
  scale_fill_manual(
    labels = c("Other", "Falls"), 
    values = c("#619cff", "#f8766d")) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
age_fall_proportion_plot

# ggsave("deaths_age_proportion_plot.png", age_fall_proportion_plot,
#        width = 1500, height = 1200, units = "px")
```

## Logistic Regression

### Training

Fit a logistic regression model using age as a predictor.

```{r}
######################
# ACTIVITY: you can try different predictors to build different models. Update the equation (falls ~ age_group_num)
######################

deaths_falls_logistic_1d <- glm(
  falls ~ age_group_num,
  data   = deaths_patient_falls_train,
  family = "binomial")
```

### Validation

```{r}
deaths_falls_1d_valid_pred <- predict(deaths_falls_logistic_1d, 
                                      newdata = deaths_patient_falls_valid)
```

Plot to see the model

```{r}
logistic_1d_model_df <- data.frame(
  x = 0:100,
  y = 1/(1+ exp(-1 * (0:100 * deaths_falls_logistic_1d$coefficients[2] + deaths_falls_logistic_1d$coefficients[1] ) ))
  )

logistic_1d_fit_plot <- ggplot(deaths_patient_falls_train,
       aes(x = age_group_num,
           y = as.numeric(falls))) + 
  geom_jitter(alpha  = 0.01, 
              height = 0.1, 
              width  = 2) +
  geom_line(data = logistic_1d_model_df,
            aes(
              x = x,
              y = y),
            col = "red") + 
  labs(
    x        = "Age",
    y        = "Death by Falls (0=False, 1=True)",
    title    = "Logistic Regression Model",
    subtitle = "risk ~ age"
  ) +
  theme_light()
logistic_1d_fit_plot

# ggsave("deaths_logistic_1d_fit_plot.png", logistic_1d_fit_plot,
#        width = 1500, height = 1200, units = "px")
```

Plot ROC curve
```{r}
deaths_falls_1d_valid_roc <- roc.curve(
  deaths_falls_1d_valid_pred[deaths_patient_falls_valid$injury_type == "Falls"],
  deaths_falls_1d_valid_pred[deaths_patient_falls_valid$injury_type != "Falls"], 
  curve = TRUE)

plot(deaths_falls_1d_valid_roc)
```

Plot balance of sensitivity and specificity
```{r}
# Calculate optimal threshold using GMA
deaths_falls_1d_valid_opt <- best_gma(deaths_falls_1d_valid_roc)

# calculate probability using sigmoid to replace threshold
deaths_falls_1d_valid_roc_prob           <- deaths_falls_1d_valid_roc
deaths_falls_1d_valid_roc_prob$curve[,3] <- 1/(1 + exp(-1 * deaths_falls_1d_valid_roc_prob$curve[,3]))
deaths_fall_1d_valid_spe_sen_plot        <- plot_spe_sen(deaths_falls_1d_valid_roc_prob)

deaths_fall_1d_valid_spe_sen_plot <- deaths_fall_1d_valid_spe_sen_plot + 
  geom_vline(aes(xintercept = 1/(1 + exp(-1 * deaths_falls_1d_valid_opt$opt_thresh))),
             linetype = 2) +
  annotate("text", 
           x     = 1/(1 + exp(-1 * deaths_falls_1d_valid_opt$opt_thresh)),
           y     = 0.1,
           hjust = 1.1,
           label = "Optimal Threshold") +
  labs(
    x        = "Threshold",
    y        = "Value",
    col      = "Metric",
    title    = "Sensitivity and Specificity at Different Thresholds",
    subtitle = paste0("Optimal Threshold = ", round(1/(1 + exp(-1 * deaths_falls_1d_valid_opt$opt_thresh)), 2),
                      ", Sensitivity = ", round(deaths_falls_1d_valid_opt$opt_sen, 2),
                      ", Specificity = ", round(deaths_falls_1d_valid_opt$opt_spe, 2))
  )
deaths_fall_1d_valid_spe_sen_plot

# ggsave("deaths_fall_1d_valid_opt_thresh_plot.png", deaths_fall_1d_valid_spe_sen_plot,
#        width = 1500, height = 1200, units = "px")
```

```{r}
# add line to denote threshold
risk_cutoff <- 1/(1 + exp(-1 * deaths_falls_1d_valid_opt$opt_thresh))
age_cut_off <- (deaths_falls_1d_valid_opt$opt_thresh -
                  deaths_falls_logistic_1d$coefficients[1])/
  deaths_falls_logistic_1d$coefficients[2]

logistic_1d_fit_thresh_plot <- logistic_1d_fit_plot +
  geom_vline(aes(xintercept = age_cut_off),
             linetype = 2) +
  geom_segment(
    aes(x    = -Inf, 
        xend = age_cut_off,
        y    = risk_cutoff, 
        yend = risk_cutoff),
    linetype = 2) +
  annotate("text",
           x     = age_cut_off,
           y     = 0.5,
           hjust = -0.1,
           label = "Predict Death by Fall") +
  annotate("text",
           x     = age_cut_off,
           y     = 0.5,
           hjust = 1.1,
           label = "Predict Death by Other Injury")
logistic_1d_fit_thresh_plot

# ggsave("deaths_logistic_1d_fit_plot.png", logistic_1d_fit_thresh_plot,
#        width = 1500, height = 1200, units = "px")
```

Explore different thresholds
```{r}
######################
# ACTIVITY: Try other thresholds (between 0 and 1) other than 0.5 to see how the predictions change.
######################

deaths_falls_1d_valid_class <- 1/(1 + exp(-1 * deaths_falls_1d_valid_pred)) >= 0.5

head(data.frame(
  pred         = 1/(1 + exp(-1 * deaths_falls_1d_valid_pred)),
  pred_class   = deaths_falls_1d_valid_class, 
  actual_class = deaths_patient_falls_valid$falls,
  age          = deaths_patient_falls_valid$age_group))

# confusion matrix
table(deaths_falls_1d_valid_class, deaths_patient_falls_valid$falls)
```


### Testing
```{r}
deaths_falls_1d_test_pred <- predict(deaths_falls_logistic_1d, 
                                     newdata = deaths_patient_falls_test)
deaths_falls_1d_test_roc <- roc.curve(
  deaths_falls_1d_test_pred[deaths_patient_falls_test$injury_type == "Falls"],
  deaths_falls_1d_test_pred[deaths_patient_falls_test$injury_type != "Falls"],
  curve = TRUE)
deaths_falls_1d_test_class <- deaths_falls_1d_test_pred >= deaths_falls_1d_valid_opt$opt_thresh


table(deaths_falls_1d_test_class, deaths_patient_falls_test$falls)

# calculate specificity and sensitivity
deaths_falls_1d_test_thresh_index <- which(deaths_falls_1d_test_roc$curve[,3] < deaths_falls_1d_valid_opt$opt_thresh)[1]
deaths_falls_1d_test_spe          <- 1 - deaths_falls_1d_test_roc$curve[deaths_falls_1d_test_thresh_index, 1]
deaths_falls_1d_test_sen          <- deaths_falls_1d_test_roc$curve[deaths_falls_1d_test_thresh_index, 2]
```

Sensitivity on the test set is `r deaths_falls_1d_test_sen`. Specificity on the test set is `r deaths_falls_1d_test_spe`.